Missing data: a simple introduction

### 2020-03-03 Draft1.tex
Finished a first draft of all major theory sections

	\subsection{What is imputation?}
	Imputation refers to substituting or ''filling in'' of missing values with plausible estimates.The main strength of imputation is that it does not change the nature of the subsequent analysis. It makes it much easier for the data processor and analyst to work together since the processor's techniques won't interfere with the analysts'. This is, at least, the goal of imputation. In certain situation the information available to the imputer might differ to that available to the analyst. In such cases difficulties arise. The concept of congeniality, which we will define shortly, helps us overcome some of these difficulties. Another very practical benefit, and a strong argument for its use, is the performance gains to be had from using it. This will become clear in the experiments section.

@@

	\subsection{What is valid inference?}
	Techniques are based on the bayesian paradigm, but inference is often done in the frequentist one.

@@

	MI is motivated under the Bayesian paradigm. This causes some (Reference Fay and stuff from MENG) to question whether inference under MI is valid at all. To address this Meng [1994] developed the concept of congeniality, which we will go into more detail on soon.

	\subsection{Congeniality: when the imputer's and analysts models differ}
	As mentioned in the introduction, modern data analysis' multiple phases mean that the assumptions underlying each phase and the relationship between the phases is an important object of study. Meng's [meng 1994] concept of congeniality provides a theoretical framework from which to evaluate these differences and most of this section is derived from his paper.

@@

	Additionally, let $g(Y^{m}|{Y^{o}, M}, A)$ be the imputation model, where A represents possible additional data the imputer has access to. Meng continues by stating ''The analysis procedure $P$ is said to be congenial to the imputation model $g(Y^{m}|{Y^{o}, M}, A)$ if one can find an f such that 1) f is congenial to $P$ and 2) the posterior predictive density for $Y^{m}$ derivd under f is identical to the imputation model $$f(Y{m}|{Y^{o}, M}) = g(Y{m}|{Y^{o}, M}, A) \forall Y^{m}$$

@@

	''When the imputation model class and the (embedded) analyst’s model class
	differ, the behavior of Rubin’s rules becomes very complicated, capable of producing inconsistent variance estimators, a matter that has received recurrent
	criticisms'' [Meng 2017]

	Congeniality is an important concept in imputation and has a big impact on the statistical validity of MI. In short, uncongeniality arises when the analysis procedure differs substantially from the imputation model. There are many reasons for such difference, for example, the imputer might wish to include as many variables as possible to achieve an accurate imputation, while the analyst might be interested in only a subset of the covariates and the relationship between these. However, as Meng [1994] states, ''If the imputer's model is reasonably accu-
	rate, then following the multiple-imputation recipe
	prevents the analyst from producing inferences with
	serious nonresponse biases''

@@

	\textbf{Single imputation}
	Single imputation is the building block of MI, and .. Single imputation can be done in a variety of ways, but the most straightforward and easy to understand is the following: create a joint posterior distribution ... Draw from this distribution to fill in your dataset, and then do your analysis.

	\section{Multiple imputation}
	One sizeable weakness of the earlier imputation methods is that they use points as estimates of distributions. Hence, subsequent measures of precision are biased downwards. Multiple imputation solves this problem by creating multiple imputed datasets drawn from a posterior distribution and then combines them in the final step. This preserves the distributional variation in the missing data, and ... Increases in available computation power has made MI a much more practical method to implement, as one of the weaknesses that earlier impeded its adoption was that it was computationally expensive.

	Multiple imputation can be thought of as three distinct steps:
	1) M imputed datasets are generated through repeated random draws from the predictive distribution of the missing values under a particular model for missingness.
	2) The intended analysis is performed separately on all M datasets
	3) The results of the analysis are combined using Rubin's or Meng's rules to create a single MI result.

	Between imputation variance and within imputation variance

	It should be noted that even though there is no strict mathematical proof that MI yields valid inference, the empirical evidence presents a strong case that MI is a useful tool. Comparing it to the other methods available, one finds that alternatives are on the whole less effective.

@@

	We often want to include as many covariates as possible. X have shown that even mildly correlated auxiliary variables can improve subsequent inference [cite annotadet article]. However, this can cause the imputation and and analyses models to be uncongenial.

	\subsection{Rubin's rules and Meng's adaptations}

	At step 2) of MI, we have M sets of parameter estimates and their variances $\hat{\theta}_{i},\hat{Var_{i}} \qquad i=0,...,M$. The next step is to combine these into a single estimate. Under congeniality Rubin's combining rules suffice for giving us statistically valid estimates. The parameter estimate is simply the average $$\bar{\theta} = \frac{1}{M}\sum_{i=0}^{M}\hat{\theta}_{i}$$ and the variance associated with $\bar{\theta}$ is given by $$T = \bar{Var}+(1+\frac{1}{M}B)$$ where $$\bar{Var} = \frac{1}{M}\sum_{i=0}^{M}\hat{Var}_{i}$$ measures the within imputation variability and $$B = \frac{1}{M-1}\sum_{i=0}^{M}(\hat{\theta_{i}}-\bar{\theta})(\hat{\theta_{i}}-\bar{\theta})^{T}$$ measures the between imputation variability. The adjustment $(1+\frac{1}{M})$ is due to the finite number of imputations. If we let $M\rightarrow\infty$ then...

	If models are uncongenial, we use Meng's rules.

@@

@@

@@

@@

	\subsection{TODO}
	Difference between baysian and frequentist paradigm.
	Citation example:
	\cite[P.10]{grund_pooling_2016}.

>>>>

	\subsection{What is imputation?}
	Imputation refers to substituting or ''filling in'' of missing values with plausible estimates. Often the person creating or maintaining a dataset will have access to more information than the one analysing it. Using this information in correcting for missing data means that the imputer can do a better job than the analyst. The main strength of imputation is that it does not change the nature of the subsequent analysis. It makes it much easier for the data processor and analyst to work together since the processor's techniques won't interfere with the analysts'. This is, at least, the goal of imputation. In certain situation the information available to the imputer might differ to that available to the analyst. In such cases difficulties arise. The concept of congeniality, which we will define shortly, helps us overcome some of these difficulties. Another very practical benefit, and a strong argument for its use, is the performance gains to be had from using it. This will become clear in the experiments section.

@@

	\subsection{What is valid inference?}
	Some have questioned whether methods developed under the Bayesian paradigm leads to valid frequentist inference (Meng1994). However, Xia and Meng(2017), Meng(1994), and Rubin(1996) have shown that such Bayesian models yield valid inference in the asymptotic case, and even in the non-asymptotic case following the below definition given certain conditions which we will explore later.

@@

	\subsection{Congeniality: when the imputer's and analysts models differ}
	As mentioned in the introduction, modern data analysis' multiple phases mean that the assumptions underlying each phase and the relationship between the phases is an important object of study. Differences might cause the subsequent analysis to be biased, or in the worst case, invalid. Meng's [meng 1994] concept of congeniality provides a theoretical framework from which to evaluate these differences and most of this section is derived from his paper.

@@

	Additionally, let $g(Y^{m}|{Y^{o}, M}, A)$ be the imputation model, where A represents possible additional data the imputer has access to. Meng continues by stating ''The analysis procedure $P$ is said to be congenial to the imputation model $g(Y^{m}|{Y^{o}, M}, A)$ if one can find an f such that 1) f is congenial to $P$ and 2) the posterior predictive density for $Y^{m}$ derived under f is identical to the imputation model $$f(Y{m}|{Y^{o}, M}) = g(Y{m}|{Y^{o}, M}, A) \forall Y^{m}$$

@@

	Congeniality is an important concept in imputation and has a big impact on the statistical validity of MI. In short, uncongeniality arises when the analysis procedure differs substantially from the imputation model. There are many reasons for such difference, for example, the imputer might wish to include as many variables as possible to achieve an accurate imputation, while the analyst might be interested in only a subset of the covariates and the relationship between these. However, as Meng [1994] states, ''If the imputer's model is reasonably accu-
	rate, then following the multiple-imputation recipe
	prevents the analyst from producing inferences with
	serious nonresponse biases''. Meng goes on to show that even with an uncongenial imputation model, inferences performed on a MI dataset still yield better inference than the best possible incomplete data inference.

@@

	\textbf{Single imputation}
	Single imputation is the building block of MI, and many of it's deficiencies are corrected for when expanded to MI. It can be done in a variety of ways, but the most straightforward and easy to understand is the following: create a joint posterior distribution of your missing data given your observed data, i.e. $f(Y^{m}|Y^{o}, M)$. Draw from this distribution to fill in your dataset, then perform your desired analysis. How you specify the missing data model can determine the efficacy of this method, but it is otherwise uncomplicated. We will therefore focus the majority of our analysis on the MI case.

	\section{Multiple imputation}
	One sizeable weakness of the earlier imputation methods is that they use points as estimates of distributions. Hence, subsequent measures of precision are biased downwards. Multiple imputation solves this problem by creating multiple imputed datasets drawn from a posterior distribution and then combines them in the final step. This preserves the distributional variation in the missing data, and leads to better variance estimates. Increases in available computation power has made MI a much more practical method to implement, as one of the weaknesses that earlier impeded its adoption was that it was computationally expensive. Many of the reservations people had with MI regarding inference concerned MI with few imputed datasets M, typically in the 5-10 range. With larger M and more advanced theoretical framework for combining the resulting analyses, one can achieve strong results in a large number of situations, see Xia and Meng (2017), Meng (1994), Rubin (1996), and Madley-Dowd et al. (2019).

	Multiple imputation can be thought of as three distinct steps:
	1) M imputed datasets are generated through repeated random draws from the predictive distribution of the missing values ($f(Y^{m}|Y^{o}, M)$) under a particular model for missingness.
	2) The intended analysis is performed separately on all M datasets
	3) The results of the analysis are combined using Rubin's or Meng's rules to create a single MI result.

	It should be noted that even though there is no strict mathematical proof that MI yields valid inference in the case of finite M, the empirical evidence presents a strong case that MI is a useful tool. Comparing it to the other methods available, one finds that alternatives are on the whole less effective. %Additionally, Meng (1994) shows us that for sufficiently many imputations we have that the variance is less than for one calculated in the complete case.

	%If MI is used, the imputer should convey as clearly as possible to the analyst the form of their model and its underlying assumptions.

@@

	To create an accurate model for the missing data, we often want to include as many covariates as possible. Madley-Dowd et al. (2019) have shown that even mildly correlated auxiliary variables can improve subsequent inference. This leads us to make repeated random draws from $f(Y^{m}|Y^{o}, M, A)$, where A represents auxiliary variables. Whether or not to include auxialiary variables is a judgement best made by the imputer, as it can both be a cause and a cure for uncongeniality. Meng (1994) cautions against the use of less general imputation models (e.g. ignoring covariates/auxiliary variables).

	\subsection{Rubin's rules}

	At step 2) of MI, we have M sets of parameter estimates and their variances $\hat{\theta}_{i},\hat{Var_{i}} \qquad i=0,...,M$. The next step is to combine these into a single estimate. Under congeniality Rubin's combining rules suffice for giving us statistically valid estimates. The parameter estimate is simply the average $$\bar{\theta} = \frac{1}{M}\sum_{i=0}^{M}\hat{\theta}_{i}$$ and the variance associated with $\bar{\theta}$ is given by $$T = \bar{Var}+(1+\frac{1}{M}B)$$ where $$\bar{Var} = \frac{1}{M}\sum_{i=0}^{M}\hat{Var}_{i}$$ measures the within imputation variability and $$B = \frac{1}{M-1}\sum_{i=0}^{M}(\hat{\theta_{i}}-\bar{\theta})(\hat{\theta_{i}}-\bar{\theta})^{T}$$ measures the between imputation variability. The adjustment $(1+\frac{1}{M})$ is due to the finite number of imputations. If we let $M\rightarrow\infty$ then we achieve asymptotic equality to the true variance estimate.

	Xia and Meng (2017) state that ''When the imputation model class and the (embedded) analyst’s model class
	differ, the behavior of Rubin’s rules becomes very complicated, capable of producing inconsistent variance estimators, a matter that has received recurrent
	criticisms''. They argue that in the uncongenial case , which is often the rule rather than the exception, doubling Rubin's variance measure will never underestimate the true variance. This holds for any degree of uncongeniality, and provides a surprisingly simple solution to the problem of uncongeniality. For Univariate data, Xia and Meng (2017) suggest replacing $T$ with $\tilde{T}= (\sqrt{\bar{Var}}+\sqrt{B})^{2}+\frac{1}{M}B_$. These methods may be viewed as highly conservative, but in the context of MI where uncongeniality between analysts and imputer is unknown and potentially large, it pays to be as insured as possible against invalid inference. However, even if these methods are not used and the models are uncongenial, we note that Meng (1994) has shown that MI is still superior to incomplete data procedures.

@@

	%Improper specification of the imputation model causes uncongeniality and introduces bias in the data that can distort inference.

@@

	%Needs more pitfalls and criticism

@@

	%Needs to be more comprehensive.

@@

	\subsection{TODO}
	Difference between baysian and frequentist paradigm.
	Citation example:
	Self efficiency.

	\subsection{Final notes}
	This paper has gone into great detail on MI and missing data in general. However, a rigorous comparison between likelihood based methods and MI has not been performed in the literature to the same extent as this paper. Espceially in the case of uncongenial imputation models, one would think likelihood based methods might show some promise.
	\cite[P.10]{grund_pooling_2016}.
